\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{algorithm} 
\usepackage{algpseudocode} 
\modulolinenumbers[5]


\begin{document}

\begin{frontmatter}

\title{Efficient n-gram, skipgram and flexgram modelling with Colibri Core}
\author{Maarten van Gompel \\ Centre for Language Studies \\ Radboud University Nijmegen}

\end{frontmatter}

\linenumbers

\begin{abstract}
Counting n-grams lies at the core of any frequentist corpus analysis and is
often considered a trivial matter. Going beyond consecutive n-grams to patterns
such as skipgrams and flexgrams increases the demand for efficient
solutions. The need to operate on big corpus data does so even more.  Lossless
compression and non-trivial algorithms are required to lower the memory demands,
yet retain good speed. Colibri Core is software
for the efficient computation of n-grams, skipgrams and flexgrams from corpus
data. The resulting pattern models can be analyzed and compared in various
ways. The software offers a programming library for C++ and
Python, as well as command-line tools.
\end{abstract}

\section{Introduction}

The n-gram, a sequence of $n$ consecutive word tokens, is a core concept for
many Natural Language Processing (NLP) applications. One of the most basic NLP
tasks is to read corpus text and compute an $n$-gram frequency list, elementary
for any kind of statistical analysis. The unigram frequency list, i.e.\ the word
frequency list, is the simplest instance of this task which is especially
ubiquitous. Computing $n$-gram frequency on a corpus text is fairly trivial,
and any beginning computer science student will have no trouble to accomplish
this in just a few lines of code in a modern high-level programming language.
However, Optimising this to reduce memory constraints, improve speed, and scale to large
data, is a more complex matter. Colibri Core, the NLP software we introduce here, offers 
efficient algorithms to do this.

N-grams are typically distributed in a Zipfian fashion, implying there are only
a few high-frequency patterns, with words such as common function words in the
lead, and there is a long tail of patterns that occur only very sparsely. This
basic fact makes counting a notoriously memory-hungry enterprise, as patterns
occurring below a minimum frequency treshold can not be discarded from
memory until the entire corpus has been processed sequentially. 

When working with large data sets and higher-order $n$-ngrams, this memory
problem becomes apparent quickly when trivial solutions are employed. Colibri
Core, on the other hand, offers tools and programming libraries that are
heavily optimised to 1) reduce memory usage, and 2) have high performance.

The task of finding $n$-grams is generalised in Colibri Core to the task of
finding \emph{patterns} or \emph{constructions} (we use the terms
interchangeably). Furthermore, once patterns are identified, resulting in a
\emph{pattern model}, Colibri Core can extract relations between the patterns.

As the name Colibri Core suggests, the software is geared to provide
\emph{core} functionality for modelling patterns and exposes this functionality
as a programming library as well as through command line tools. It aims to
provide a solid foundation upon which more specialised software can be built,
such as software for language modelling. The software is aimed at
NLP software developers and researchers with a solid technical background. 

Colibri Core provides memory-based techniques where models are held entirely in
memory to guarantee maximum performance on lookup and computation. It does
place considerable memory requirements on the system on which is it run. This
approach can be contrasted to e.g.\ database approaches which have much higher
latency.

\section{Patterns}

We distinguish three categories of patterns, and define them as follows:

\begin{enumerate}
    \item N-grams -- A sequence of $n$ word tokens that are all consecutive.
    \item Skipgrams -- A fixed-length sequence of $p$ word tokens and $q$ token
        placeholders/wildcards with total length $n$ ($n=p+q$), the placeholders constitute gaps or skips and a skipgram can contain multiple of these. In turn, a gap can span one or more tokens. 
    \item Flexgrams -- A sequence with one or more gaps of variable length, which implies the pattern by itself is of undefined length.
\end{enumerate}

Our definitions are defined narrowly and, with exception of $n$-grams do not
necessarily match up precisely to the way the concepts are used in other studies. Some
may use the term skipgram to include what we call flexgram.  The term
``elastigram'' is sometimes also seen in the literature to refer to flexgrams. 

Skipgrams are used in the field to obtain a higher level of generalisation than
can be offered by n-grams. They can, for instance, be found as features in
classification tasks \citep{DHONDT}, or as a means in language modelling to
decrease perplexity \citep{Guthrie06}.

Dealing with word tokens implies that the corpus data has to be in a
tokenised form. We start from the basis of plain-text corpus data, containing one
``unit'' per line; a unit can correspond to a sentence, paragraph, tweet,
or whatever unit is deemed appropriate for the task at hand. Corpus data can
alternatively be provided in FoLiA XML format \cite{FOLIAPAPER} as well, although linguistic
annotation will be ignored.

Text data is typically stored as a string of characters. The characters
themselves draw their denotation from a character encoding. The storage of a
huge amount of strings is inefficient from a memory perspective, considering
the fact that words follow a Zipfian distribution. Colibri Core therefore works
on the basis of a lossless compression, in which each unique word token is
assigned a numeric type identifier, which we call a \emph{class}. This
effectively defines the ``vocabulary'' of the data, which we call a
\emph{class encoding}. A pattern is not represented as an array of
characters, but as an array of these classes instead. Further lossless
compression is achieved by holding this array of classes in a dynamic-length
byte representation, in which low class values can be stored in fewer bytes
than high class values. Classes $0$ to $127$ can be stored in a single byte,
higher classes require at least two bytes. To achieve maximum compression,
classes are assigned to word tokens based on frequency (i.e.\ entropy
encoding): words with the highest frequency receive the lowest classes.  This
is essentially a variant of Huffman coding \citep{HUFFMAN}. Some of the lowest
classes are reserved for special purposes, e.g. to delimit sentences (class 0) or as
markers for out-of-vocabulary words (class 2) or skipped content (classes 3 and 4).

Of each byte in the class representation, the first bit is reserved as a continuation
marker. As long as the continuation marker is high, the next byte is part of
the class. When it is low, we know we are at the final byte of a class
representation. The class itself is stored in the remaining 7-bits of each
byte. In practise this results in good compression and reduces memory usage.

To encode a text corpus, a class encoding needs to be computed first. To decode
the encoded corpus back to plain text, the class encoding is needed again.
Colibri Core provides tools and exposes library functions to do this.

\section{Informed Iterative Counting}

N-gram frequency lists are often parametrised by a certain threshold. $N$-grams
with a frequency below threshold are pruned. Because of the Zipfian nature of
our $n$-gram distribution, we can circumvent the problem of holding a huge
amount of patterns in memory that do not meet the threshold. We do this by
employing informed counting. Informed counting is an iterative procedure, shown
in pseudo code in Algorithm~\ref{alg:ngramcounting}. Here we take $m$ to be the
maximum $n$-gram order we intend to extract, the whole corpus is then processed
for each $n$ where $1<n\leq m$, extracting the respective $n$-grams at each
iteration. This means that at each iteration we can consult the results of the
previous iteration. We can then readily discard any $n$-gram with $n>1$ for
which either of the two $n-1$-grams that it contains does not meet the
threshold, as it follows that the $n$-gram can never meet the threshold either.
By outright discarding an $n$-gram we do not need to store it and its count in
memory. After each iteration of $n$, we prune all the $n$-grams that did not
reach the threshold.


\begin{algorithm} \caption{Informed Iterative Counting for n-grams.  Take $m$
to be the maximum $n$-gram order we intend to extract, $t$ to be the minimum occurrence threshold, and $M$ to be the
pattern model in memory, with unigrams already counted in the more trivial fashion.}
\label{alg:ngramcounting}
\begin{algorithmic}
\For {$n \in 2..m$}
    \For {$line \in corpus$}
        \For {$ngram \in extractngrams(line,n)$}
            \State  $nm1gram_1, nm1gram_2 \leftarrow extractngrams(ngram,n-1)$
            \If {$M(nm1gram_1) \geq t$ \& $M(nm1gram_2) \geq t$}
                \State $M(ngram) \leftarrow M(ngram) + 1$
            \EndIf
        \EndFor 
    \EndFor
    \State $M \leftarrow prunengrams(M,n,t)$
\EndFor \\
\Return{M}
\end{algorithmic}
\end{algorithm}

Though not expressed in the simplified algorithm above, the actual
implementation accounts for more parameters, such as setting a lower bound to
$n$. The amount of back-off, going all the way up to $m-1$ here, can also be
fine-tuned.

A performance evaluation of this algorithm has been performed by comparing a
naive implementation in Python, a simple implementation of informed iterative
counting in Python, and the optimised C++ implementation in Colibri Core. To
this end, we use a corpus of Dutch translations of TED talks of $127,806$
sentences and $2,330,075$ words\footnote{The data is from the IWSLT 2012
Evaluation Campaign,
\url{http://hltc.cs.ust.hk/iwslt/index.php/evaluation-campaign/ted-task.html\#MTtrack}.
Tokenisation was performed using ucto, \url{http://ilk.uvt.nl/ucto}.}. We set
the occurrence threshold to $2$ and extract everything from unigrams to
$8$-grams. The naive implementation, before pruning, consumes $2737$ MB and
runs in $23$ seconds\footnote{All these experiments were conducted on a system with
an Intel Core i7-4770K CPU \@ 3.50GHz, 16GB RAM, running Linux kernel 4.1}. A Python implementation of informed iterative counting
significantly lowers memory usage to $167$ MB, at the cost of increased CPU
time; $65$ seconds. The Colibri Core implementation runs in $9.5$ seconds and
takes memory down to just $76$ MB. These gains can be attributed to the class
encoding and the fact that Colibri Core is implemented in C++.

\subsection{Informed Skipgram Counting}
\label{sec:skipgramcount}

The computation of skipgrams is parametrised by an upper limit $l\leq m$ in the number of
tokens, as the possible configuration of gaps increases exponentially with the
total length spanned. It first requires a count of all $n$-grams where $0<n\geq l$. 

The algorithm, shown in Algorithm~\ref{alg:skipgramcount} considers all
possible ways skips can be inserted in all of the n-grams in the model. It can
discard a skipgram candidate by looking at the non-skipped parts that make up the
skipgram, and checking whether those exceed the set threshold. 

\begin{algorithm} \caption{Informed Counting for skipgrams.  Take $l$
to be the maximum skipgram order we intend to extract, $t$ to be the minimum occurrence threshold, and $M$ to be the
pattern model in memory, with ngrams already counted.}
\label{alg:skipgramcount}
\begin{algorithmic}
\For {$n \in 3..l$}
    \For {$ngram \in getngrams(M,n,t)$}
        \For {$skipgram \in possibleconfigurations(ngram)$}
            \State $docount \leftarrow True$
            \For {$part \in parts(skipgram)$}
            \If {$M(part) < t$}
                    \State $docount \leftarrow False$
                    Break
                \EndIf
            \EndFor 
            \If {$docount$}
                \State $M(skipgram) \leftarrow M(skipgram) + 1$
            \EndIf
        \EndFor 
    \EndFor
    \State $M \leftarrow pruneskipgrams(M,n,t)$
\EndFor \\
\Return{M}
\end{algorithmic}
\end{algorithm}

In this algorithm, the \texttt{possibleconfigurations(ngram)} function returns
all possible skipgram configurations for the given $n$-gram. Note that
the configuration of gaps depends only on the length of the $n$-gram, regardless
of its content, and can therefore easily be pre-computed. The
\texttt{parts(skipgram)} function returns all consecutive parts that are
subsumed in the skipgram.

Like Algorithm~\ref{alg:ngramcounting}, Algorithm~\ref{alg:skipgramcount}
assumes a threshold $t>1$. When $t=1$, more trivial algorithms are
invoked, as the user does not want to prune anything. These make only a single
pass over the data. For skipgrams this leads to an explosion of
resulting patterns, exponential with number of tokens, and is best avoided. The
default will be $t=2$ in most cases.

\subsection{What counts?}
\label{sec:whatcounts}

The counting algorithms are parametrised by a large number of parameters which
are not shown in Algorithm~\ref{alg:ngramcounting} and
Algorithm~\ref{alg:skipgramcount} to reduce complexity. The set of
parameters allows the user to influence precisely what is counted. This is one
of the main assets of Colibri Core. Parameters exist to affect the following:

\begin{itemize}
    \item The minimum and maximum length (in words/tokens) of the n-grams
        and/or skipgrams to be extracted. Setting minimum and maximum length to
        the same value will produce a model of homogenous pattern length (e.g.
        only trigrams or words).
    \item A secondary word occurrence threshold can be configured. This is a value set higher than
        the primary occurrence threshold. Only patterns
        occuring above the primary threshold, and for which each of the
        individual words in the pattern passes the secondary threshold as well, will
        be included in the model.
    \item N-grams that are not subsumed by higher order n-grams, i.e.\ which do
        not  occur as part of a higher order n-gram in the data/model, can be pruned
        from the model. 
    \item Skipgrams can be constrained using the \emph{skip type threshold}. This
        requires that at least this number of distinct patterns must fit in any 
        skip in the skipgram. Higher values will produce fewer skipgrams, but
        typically more generic ones. 
    \item Skipgrams and n-grams are typically computed using the same
        occurrence threshold, but it is also possible to use a different threshold
        for skipgrams.
\end{itemize}

\section{Pattern Models}

The pattern model is a $key \mapsto value$ store, where the keys correspond to
patterns. The underlying C++ library allows for choosing the actual underlying
container implementation through templating. The default container datatype is
a hash map\footnote{using the \texttt{unordered\_map} STL container in C++11},
which guarantees $O(1)$ access and update times under ideal hashing conditions. The
hash\footnote{Spooky Hash v2 is used for hashing:
http://burtleburtle.net/bob/hash/spooky.html} is computed directly from the
binary representation of a pattern.

The use of hash maps can be contrasted to the use of suffix (or prefix) tries,
a common datastructure for storing n-grams. Although suffix tries benefit from
decreased memory use due to less overlap in pattern data. The strongly linked
nature of tries causes a significant overhead in memory use\footnote{Each pointer
consuming 8 bytes on 64-bit architectures, and one would be needed between
every two tokens.} that well exceeds the memory footprint of hash maps. For
this reason, hashmaps are the default and tries are currently not implemented
in Colibi Core. The templating, however, easily allows for such an
implementation to be added in the future.

At this point, we need to address suffix arrays \citep{Manber90} as well, which
are derived from suffix tries but with significantly decreased space
requirements. Suffix arrays with longest common prefix (LCP) arrays will
consume less memory than our hash maps, but are typically much slower to
construct and query. Though no exhaustive experiment was conducted to this
end, we did compare a predecessor of Colibri Core with a suffix array implementation
\citep{Stehouwer10} and found our implementation to be significantly faster in model
construction.

We distinguish two types of pattern models, depending on the type of the values,
which in the underlying C++ implementation is subject to templating as well:

\begin{enumerate}
 \item \textbf{Unindexed Pattern Models} -- Values are simple integers.
 \item \textbf{Indexed Pattern Models} -- Values are arrays of indices where
     the pattern's occurrences in the corpus are stored.
\end{enumerate}

Obviously, indexed pattern models make a considerably higher demand on memory.
They do, however, allow for a broader range of computations, as shall become
apparent in subsequent sections.

\subsection{Two-step training}

Training indexed patterns models is more memory intensive than training
unindexed models, especially in very large corpora (e.g. over 100 million
words). To lower the demand on memory for such corpora, we implement a
\emph{two-step training} procedure. This involves first constructing an
unindexed pattern model and subsequently constructing an indexed model on the
basis of that, by making another pass over the corpus and gathering all
indices. The gain here is in avoiding temporary storage of the indices that
will not pass the occurrence threshold but that cannot be ruled out a priori by the
informed counting algorithm.  This conservation of memory comes at the cost of
a extended execution time.

\section{Corpus Comparison}

The computation of pattern models on two or more distinct corpora, provided the
class encoding is the same for all of them, provides a basis for comparative
corpus analysis. One measure for corpus comparison introduced in the software
is the notion of ``coverage''. This metric is expressed as the number of tokens
in the test corpus that is covered by the patterns from the training corpus. It
is therefore not a commutative metric and explicitly depends on choosing a
training and a test corpus. The metric can be represented either in absolute
counts, or in normalised form as a fraction of the total amount of tokens in
the test corpus. 

To perform such comparisons, we first compute a pattern model on the training
corpus, and subsequently compute a second pattern model on the test corpus, but
\emph{constrained} by the former pattern model. The ability to train
constrained models is present throughout the software and can for instance also
be used to train a pattern model based on a custom preset list of patterns. The
previously described two-step training algorithm is also an example of
constrained training.

Summarised statistics are computed at multiple levels. Measures such as
occurrence count and coverage can be consulted for aggregates of n-grams,
skipgrams, or flexgrams. The former two can be inspected specifically for each
of the different pattern sizes present in the model, i.e.\ for each value of $n$.

The coverage metric is a fairly crude metric of corpus overlap, despite the
ability to assess it for different aggregates. A more widely established metric
for corpus comparison is log-likelihood. Log likelihood expresses how much more
likely any given pattern is for either of the two models. It therefore allows
you to identify how indicative a pattern is for a particular corpus. Our
implementation follows the methodology of~\cite{Rayson00comparingcorpora}.

\section{Relations between Patterns}

Various relations can be extracted between patterns in a pattern model, either
through the API or a dedicated query tool. For all but the first of the
relation types an indexed pattern model is required. 

\begin{itemize}
 \item \textbf{Subsumption Relation} -- $n$-gram $x y z$ subsumes $n-1$-grams $x y$ and $y z$. 
 \item \textbf{Succession Relation} -- Patterns that occur in a sequence in the corpus data. 
 \item \textbf{Instantiation Relation} -- Skipgrams or flexgrams may be
     \emph{instantiated} by other patterns. For example, ``to be \_ not \_'' be
     is instantiated by ``or \_ to'', resulting in the 6-gram ``to be or not to be''. This type of relation thus allows you to precisely determine what patterns occur in certain gaps.
 \item \textbf{Co-occurrence Relation} -- Different patterns can naturally co-occur multiple times
     within the the structural ``units'' you decided upon for the corpus (e.g.\ 
     sentences, paragraphs, tweets, etc). These units are newline delimited in
     your original input. The measure of such co-occurrence 
     can be expressed by established metrics such as Jaccard and (normalised) mutual
     pointwise information.
\end{itemize}

For each of these categories, the relationship is bidirectional, i.e.\ you can
query for the subsuming patterns as well as the subsumed patterns, the left
neighbours as well as the right neighbours, the instantiations as well as the
abstractions. The co-occurrence relationship is fully symmetrical. 

These latter three relationships rely on both the \emph{forward index} inherent
in an indexed pattern model, as well as the \emph{reverse index}, a function
from positions in the corpus to an array of patterns that are found at said
position. This can be automatically computed from the forward index. In the
actual implementation, the reverse index is not modelled in memory as a mapping from
positions to patterns, but as a much less memory-consuming mapping from
positions to unigrams; the patterns can be derived on-the-fly from that.

\subsection{Flexgram Counting}

Thus far we have explained the algorithms for n-gram counting and skipgram
counting, but have not yet done so for flexgrams, i.e. patterns with one or more
\emph{variable-width} gaps. The implementation allows
flexgrams to be computed in two different ways. The first is by 
extracting skipgrams first, and then abstracting flexgrams from these
skipgrams. In this case the flexgram computation is constrained by the same
maximum-size limit. The second method for flexgram extraction
proceeds through the co-occurence relation. A flexgram is formed whenever two
patterns (within the same structural unit) occur above a set threshold. The
implementation of this latter method is currently limited to flexgrams with a
single variable-width gap. This method is recommended when the user is
interested in long-distance flexgrams, whereas the abstraction method is
recommended when the user is more interested in having multiple gaps.

\section{Software specification}

Colibri Core consist of a C++ library, a Python binding\footnote{Both for
Python 2 as well as Python 3, written in Cython.} to said library and a set of
command-line tools. This allows for an interface to the sofware at various
levels, accommodating the need in different use cases and catering for
different levels of expertise. 

The Python binding ensures that the functionality of Colibri Core is available
from Python without sacrificing the great performance benefit native code
provides. Python was chosen as it is a high-level programming language in
widespread use in the scientific community, and the NLP community in
particular. It demands less expertise from the developer than C++ and is more
suitable for rapid prototyping.

From the command line, the following tools are available:

\begin{itemize}
    \item \texttt{colibri-patternmodeller} -- This is the core tool, it allows for the
        construction and inspection of pattern models.
    \item \texttt{colibri-classencode} -- For the construction of a class encoding and
        the encoding of corpus data
    \item \texttt{colibri-classdecode} -- For decoding encoded corpus data, given a class
    \item \texttt{colibri-comparemodels} -- Compares the frequency of patterns between
        two or more pattern models by computing log likelihood,.
\end{itemize}

There are also higher-level shell scripts that invoke the above tools, so basic
tasks can be accomplished more easily without exposing the low-level functionality:

\begin{itemize}
    \item \texttt{colibri-ngrams} -- Extracts n-grams of a particular size from the
        corpus text, in the order they occur, i.e.\ by moving a sliding window over the text.
    \item \texttt{colibri-freqlist} -- Extracts all n-grams from one or more corpus text files and outputs a frequency list. Also allows for the extraction of skipgrams. By default all n-grams are extract, but an occurrence threshold can be set with the -t flag.
    \item \texttt{colibri-ngramstats} -- Prints a summary report on the ngrams in one or more corpus text files. To get the full details on interpreting the output report, read the section Statistical Reports and Histograms.
    \item \texttt{colibri-histogram} -- Prints a histogram of pattern occurrence count
    \item \texttt{colibri-queryngrams} -- Interactive CLI tool allowing you to query ngrams from standard input, various statistics and relations can be outputted.
    \item \texttt{colibri-reverseindex} -- Computes and prints a reverse index for the specified corpus text file. For each token position in the corpus, it will output what patterns are found there (i.e start at that very same position)
    \item \texttt{colibri-loglikelihood} -- Computes the log-likelihood between patterns in two or more corpus text files, which allows users to determine what words or patterns are significantly more frequent in one corpus than the other.
    \item \texttt{colibri-coverage} -- Computes overlap between a training corpus and a test corpus, produces coverage metrics.
\end{itemize}

The tools and scripts in Colibri Core are explicitly geared for command-line
usage. However, a RESTful webservice as well as generic web
interface\footnote{The webservice is hosted on
\url{https://webservices-lst.science.ru.nl}} to Colibri Core is also provided using
CLAM\cite{CLAMPAPER}, to accommodate less technical end-users as well as
automated networked clients. Future work building upon Colibri Core, may focus
on offering appealing higher-level interfaces to reach a wider audience.

The software should compile and run on a wide variety of POSIX-compliant
systems, including Linux, FreeBSD and Mac OS X. Everything is released as open
source under the GNU Public License (v3)\footnote{https://gnu.org/licenses/gpl.html}.

Documentation, including API reference, is provided at
\url{http://proycon.github.io/colibri-core/doc/}. An extra tutorial for Python
is available at \url{http://proycon.github.io/colibri-core/doc/colibricore-python-tutorial.html}.

\section*{References}

\bibliographystyle{elsarticle-num}
\bibliography{coco_paper}


\end{document}
